---
title: "project"
author: "group 3-AA"
date: "2025-05-31"
output: pdf_document
---

## data clean

```{r}
library(tidyverse)
```

```{r}
df_atlas <- read_delim("AtlasPlusTableData.csv")
df_ps <- read_delim("PS-Syphilis-Rates-Women-15-44-Years-by-County-US-2023.csv")
```

```{r}
df_atlas <- df_atlas %>%
  mutate(across(where(is.character), ~na_if(., "Data not available"))) %>%
  mutate(across(where(is.character), ~na_if(., "Data suppressed"))) %>%
  drop_na()
dim(df_atlas)
```

```{r}
df_ps <- df_ps %>% 
  mutate(across(where(is.character), ~na_if(.,"Unavailable/suppressed")))%>%
  drop_na()
dim(df_ps)
```

```{r}
df_atlas_merge <- df_atlas %>%
  select(-FIPS,-Cases) %>%
  rename(
    State = Geography, 
    Rate = 'Rate per 100000'
  )
head(df_atlas_merge)

df_ps_merge <- df_ps %>%
  select(-'State FIPS',- 'County FIPS',-'Age Range',-'Sex',-'Rate Category')%>%
  rename(
    State = State,
    County = County,
    Indicator = STI,
  )
head(df_ps_merge)
```


## data combine

```{r}
df_atlas_merge <- df_atlas_merge %>%
  mutate(
    County = NA_character_,
    Rate = as.numeric(gsub(",", "", Rate)),
    Year = as.numeric(stringr::str_extract(Year, "\\d{4}"))
  ) %>%
  select(Indicator, Year, State, County, Rate)

df_ps_merge <- df_ps_merge %>%
  mutate(
    Rate = as.numeric(gsub(",", "", Rate)),
    Year = as.numeric(Year)
  ) %>%
  select(Indicator, Year, State, County, Rate)

df_combined <- bind_rows(df_atlas_merge, df_ps_merge)

dim(df_combined)
```

## P1    How have syphilis rates changed across different states from 2000 to 2023? 

```{r}
df_syphilis <- df_atlas %>%
  filter(str_detect(Indicator, "Syphilis")) %>%
  mutate(
    rate = as.numeric(str_replace(`Rate per 100000`, ",", "")),
    Year = as.numeric(str_extract(Year, "\\d{4}"))
  ) %>%
  group_by(Year, Indicator) %>%
  summarise(avg_rate = mean(rate, na.rm = TRUE), .groups = "drop")

ggplot(df_syphilis, aes(x = Year, y = avg_rate, color = Indicator)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Long-Term Trends of Syphilis in the U.S. (2000–2023)",
    x = "Year",
    y = "Average Rate per 100,000 People",
    color = "Syphilis Period"
  ) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
# ggsave("syphilis.png", width = 10, height = 6, dpi = 300)
```


## P2 How does syphilis compare to other STIs (gonorrhea and chlamydia) in terms of trends and severity from 2000 to 2023?

```{r}
df_subset <- df_combined %>%
  filter(Year %in% c(2018, 2022), is.na(County)) %>%
  filter(Indicator %in% c("Primary and Secondary Syphilis"))

ggplot(df_subset, aes(x = Rate, y = fct_reorder(State, Rate), fill = Indicator)) +
  geom_col() +
  facet_wrap(~Year) +
  labs(title = "STI Rates by State in 2018 and 2022",
       x = "Rate per 100,000", y = "State") +
  scale_fill_manual(values = c("Primary and Secondary Syphilis" = "limegreen"))+
  theme(axis.text.y = element_text(size = 4))
# ggsave("state.png", width = 10, height = 6, dpi = 300)
```

```{r}
df_national <- df_combined %>%
  filter(is.na(County)) %>%
  filter(Indicator %in% c("Chlamydia", "Gonorrhea", "Primary and Secondary Syphilis")) %>%
  group_by(Year, Indicator) %>%
  summarise(avg_rate = mean(Rate, na.rm = TRUE), .groups = "drop")

ggplot(df_national, aes(x = Year, y = avg_rate, color = Indicator)) +
  geom_line() +
  geom_point() +
  labs(
    title = "National Average STI Rates Over Time (2000–2023)",
    y = "Average Rate per 100,000", x = "Year"
  ) +
  scale_color_manual(values = c(
    "Chlamydia" = "red",
    "Gonorrhea" = "blue",
    "Primary and Secondary Syphilis" = "limegreen"
  ))
# ggsave("all.png", width = 10, height = 6, dpi = 300)
```

## P3: Do states with historically rising syphilis trends also exhibit high county-level infection rates among women aged 15–44 in 2023?

```{r}
df_syphilis <- df_combined %>%
  filter(Indicator == "Primary and Secondary Syphilis", is.na(County))

ggplot(df_syphilis, aes(x = Year, y = Rate, color = State)) +
  geom_line() +
  labs(title = "Primary and Secondary Syphilis Rates by State Over Time",
       y = "Rate per 100,000")
# ggsave("primary and secondary.png", width = 10, height = 6, dpi = 300)
```


## P4 Which states and regions should be prioritized for future funding and interventions?

```{r}
df_sd_county <- df_combined %>%
  filter(
    State == "South Dakota",
    Indicator == "Primary and secondary syphilis",
    !is.na(County)
  )

ggplot(df_sd_county, aes(x = Rate, y = fct_reorder(County, Rate))) +
  geom_col(fill = "orange") +
  labs(
    title = "Primary and Secondary Syphilis Rates by County in South Dakota (2023)",
    x = "Rate per 100,000",
    y = "County"
  ) +
  theme(axis.text.y = element_text(size = 4))
# ggsave("county.png", width = 10, height = 6, dpi = 300)
```
